---
title: "Making and plotting causal models"
format: 
   revealjs:
      embed-resources: true
      theme: serif
      slide-level: 3
      slide-number: true
      toc-depth: 2
      show-slide-number: all
      preview-links: auto
      number-sections: true
      link-color: orange
      smaller: true
---



Plotting functionality makes use of the Sugiyama layout from igraph which plots nodes to reflect their position in a causal ordering.

The `plot` method calls `plot_model` and passes provided arguments to it.

## A basic plot:

```{r bbasic, fig.cap = "Simple model"}
model <- make_model("X -> Y")

model |> plot_model()
```

## ggplot layers

The model that is produced is a `ggplot` object and additional layers can be added in the usual way.

```{r bgg, fig.cap = "Adding additional `ggplot` layers"}
model |>
  plot_model()  +
  annotate("text", x = c(1, -1) , y = c(1.5, 1.5), label = c("Some text", "Some more text")) +
  coord_flip()
```


## Adding labels

Provide labels in the same order as model nodes.

```{r blabs, fig.cap = "Adding labels"}
model <- make_model("A -> B -> C <- A")


# Check node ordering
inspect(model, "nodes")

# Provide labels
model |>
   plot_model(
     labels = c("This is A", "Here is B", "And C"),
     nodecol = "white", textcol = "black")
```

## Controlling  positions

You can manually set positions using the `x_coord` and `y_coord` arguments.

You can manually set positions using the `x_coord` and `y_coord` arguments.

```{r bcoord, fig.cap = "Specifying coordinates"}
model |>
  plot(x_coord = 0:2,  y_coord = c(0, 2, 1))
```


## Controlling  color

You can manually control node color and text color for all nodes together or separately.

```{r bcolor, fig.cap = "Controlling colors"}
model |>
  plot(x_coord = 0:2,  y_coord = c(0, 2, 1),
       nodecol = c("blue", "orange", "red"),
       textcol = c("white", "red", "blue"))
```

## Models with unobserved confounding

Unobserved confounding is represented using dashed curves.


```{r bdashed, fig.cap = "Plot showing confounding"}
make_model('X -> K -> Y <- X; X <-> Y; K <-> Y') |>   plot()
```


