---
title: "Build, update, and query causal models with `CausalQueries`: Cheat Sheet"
format:
  pdf:
    documentclass: article
    geometry: "a3paper,landscape,margin=0.5in"
    fontsize: 10pt
    number-sections: false
    toc: false
header-includes:
  - \usepackage{multicol}
  - \usepackage[scaled=0.95]{helvet}
  - \usepackage{fancyhdr}
  - \usepackage{titling}
  - \usepackage{sfmath}
  - \usepackage{tcolorbox}
  - \tcbuselibrary{skins,breakable}
  - \usepackage{fontspec}
  - \renewcommand{\familydefault}{\sfdefault}
  - \renewcommand{\rmdefault}{phv}
  - \renewcommand{\sfdefault}{phv}
  - \setmainfont{TeX Gyre Heros}
  - \setsansfont{TeX Gyre Heros}
  - \setlength{\droptitle}{-6em}
  - \pretitle{\begin{center}\Huge\bfseries}
  - \posttitle{\par\end{center}\vspace{-1.6em}}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \pagenumbering{gobble}
  - \renewcommand{\headrulewidth}{0pt}
  - \renewcommand{\footrulewidth}{0.4pt}
  - |
    \makeatletter
    \renewcommand\section{\@startsection{section}{1}{\z@}{0.6em}{0.3em}{\sffamily\bfseries\Large}}
    \makeatother
  - |
    \definecolor{CalloutBorder}{HTML}{B8B2A6}
    \definecolor{CalloutBg}{HTML}{F7F6F2}
    \definecolor{CalloutTipBg}{HTML}{EEF5F0}
    \definecolor{CalloutTipBorder}{HTML}{7A9B8C}
    \tcbset{
      colback=CalloutBg,
      colframe=CalloutBorder,
      boxrule=0.3pt,
      arc=2pt,
      left=3pt,
      right=3pt,
      top=2pt,
      bottom=2pt
    }
    \tcbset{callout-tip/.style={colback=CalloutTipBg,colframe=CalloutTipBorder}}
execute:
  echo: true
  eval: true
  warning: false
  message: false
  error: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = TRUE,
  results = "markup",
  message = FALSE,
  warning = FALSE
)
test <- FALSE
library(CausalQueries)
library(knitr)
library(tidyverse)
theme_set(
  theme_bw(base_family = "Helvetica", base_size = 9) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
      text = element_text(family = "Helvetica"),
      axis.text = element_text(size = 7),
      axis.ticks = element_line(linewidth = 0.2),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold"),
      panel.border = element_rect(color = "grey30", linewidth = 0.2, fill = NA)
    )
)
```

```{=latex}
\begin{multicols}{3}
```
## 1. Make your model

`CausalQueries` works with models that assume a causal structure over **binary nodes**.  

::: {.callout-note appearance="simple"}
### a. Make your model

Provide a causal structure connecting nodes using **dagitty** syntax.

```{r update-model, eval = test}
model_1 <- make_model("X -> Y; Z -> Y")        # simple
model_2 <- make_model("Z -> X -> Y; X <-> Y")  # with confounding
```
:::

::: {.callout-note appearance="simple"}
### b. Optionally: refine your model

#### Set restrictions *(to rule out some nodal types)*

```{r, eval = test}
# naming types
make_model("X -> M -> Y") |>
  set_restrictions(labels = list(M = c("10", "11"), Y = "00"))

# using logical statements
make_model("X -> Y") |>
  set_restrictions(statement = "Y[X=1] > Y[X=0]")

# natural language helpers
make_model("X -> Y") |> set_restrictions(decreasing("X", "Y"))
```

#### Set priors *(to make use of prior knowledge)*

```{r, eval = test}
make_model("X -> Y") |> set_priors(distribution = "jeffreys")

make_model("X -> Y") |>
  set_priors(statement = "Y[X=1] > Y[X=0]", alphas = .25)
```

#### Set parameters *(to imagine a particular world)*

```{r, eval = test}
make_model("X -> Y") |>
  set_parameters(statement  = "Y[X=1] > Y[X=0]", parameters = .5)
```
:::

::: {.callout-note appearance="simple"}
### c. Graph your model

```{r plotting, eval = FALSE}
make_model("X -> M -> Y <- X; X <-> Y") |> plot()
```

```{r , echo = FALSE, fig.height = 2, fig.width = 8}
make_model("X -> M -> Y <- X; X <-> Y") |>  plot(nodecol = "white", textcol = "black")
```

:::

::: {.callout-note appearance="simple"}
### d. Inspect your model

```{r, eval = FALSE}
make_model("X -> Y") |> inspect("nodal_types")
make_model("X -> Y") |> inspect("causal_types")
```

:::

 
```{=latex}
\columnbreak
```
## 2. Update your model

To update, pass the model and data to `update_model()`. 

::: {.callout-note appearance="simple"}
Data can be complete or missing at random. 
```{r, eval = FALSE, fig.height = 4, fig.width = 9}
model <- make_model("X -> M -> Y")
data  <- data.frame(X = 0:1, M = 1:0, Y = 0:1) |> uncount(10)

model <- update_model(model, data)
```

```{r, echo = FALSE, fig.height = 4, fig.width = 9}
model <- make_model("X -> M -> Y")
data  <- data.frame(X = 0:1, M = 1:0, Y = 0:1) |> uncount(10)

model <- update_model(model, data = data, refresh = 0)
```

Posterior distributions on parameters can be accessed directly (via `grab` or `inspect`) or when you query the model.

```{r, fig.height = 3, fig.width = 7}
model |> grab("posterior_distribution") |> 
  ggplot(aes(Y.01, Y.10)) + geom_point(alpha = .5)
```
:::

::: {.callout-tip appearance="simple"}
### Use additional Stan arguments to control updating

Other Stan arguments can be passed to `update_model()`:

- `iter` sets the number of iterations (and ultimately posterior draws)
- `chains` sets the number of chains (parallel chains can speed things up)
- many other options via `?rstan::stan`
:::

::: {.callout-tip appearance="simple"}
### Causal syntax

`CausalQueries uses` a causal syntax that lets you describe arbitrary observational or controlled values of a node.  Square brackets indicate controlled values (implied application of the “do” operator).

**An observational query:** types that produce `Y == 1`.

```{r, eval = FALSE}
make_model("X -> M -> Y <- X") |>
  get_query_types("Y==1")
```

**A simple counterfactual query:** types that produce `Y == 1` when `X` is set to `1`.

```{r, eval = FALSE}
make_model("X -> Y") |>
  get_query_types("Y[X=1]==1")
```

**A complex query:** types for which `Y == 1` when `X=1` and `M` is held constant at the value it would take if `X` were `0`.

```{r, eval = FALSE}
make_model("X -> M -> Y") |>
  get_query_types("Y[M=M[X=0], X=1]==1")
```

We provide helpers for common causal statements (`increasing`, `decreasing`, `interacts`, etc.), which are useful for querying models or setting restrictions.
:::

 
```{=latex}
\columnbreak
```
## 3. Query your model

Formulate a causal query using causal syntax and put the query to the model.


::: {.callout-note appearance="simple"}
Provide:

- causal queries (`queries`)
- conditions (`given`)
- which parameters to use (`using`)

```{r query-model}
model <- make_model("X -> M -> Y")

results <- model |>
  query_model(
    queries = list(
      ATE = "Y[X=1] - Y[X=0]",
      POS = "Y[X=1] > Y[X=0]"
    ),
    given = c(
      "All",
      "X==1 & Y==1",
      "X==1 & Y==1 & M==0"
    ),
    using = c("parameters", "priors")
  )
```
:::

::: {.callout-note appearance="simple"}
## Putting it all together

```{r full-example, eval = FALSE}

results <-
  make_model("X->M->Y") |>
  update_model(data) |>
  query_model(
    query = "Y[X=1]>Y[X=0]",
    given = c("All", "X==1 & Y==1", "X==1 & Y==1 & M==1"),
    using = c("priors", "posteriors")
  )
```

```{r, echo = FALSE}

results <-
  make_model("X->M->Y") |>
  update_model(data = data, refresh = 0) |>
  query_model(
    query = "Y[X=1]>Y[X=0]",
    given = c("All", "X==1 & Y==1", "X==1 & Y==1 & M==1"),
    using = c("priors", "posteriors")
  )
```

and view as table or plot:

```{r, eval = FALSE}
plot(results)
```

```{r, echo = FALSE}
plot(results) + theme(legend.position = "bottom")
```

:::


```{=latex}
\end{multicols}
```

::: {.callout-note appearance="simple"}
`CausalQueries` makes it easy to make, update, and query causal models defined over binary variables. Model parameters are defined using potential outcomes and updated with `rstan`. For more see [https://integrated-inferences.github.io/](https://integrated-inferences.github.io/) CC BY SA package version 1.4.4 • Updated: 2026-01-25
:::
