---
title: "Causal models for qualitative inference and mixed methods"
author: Macartan Humphreys and Alan Jacobs
format: 
  html:
    embed-resources: true
---

Helper file for [Causal models short-course](https://integrated-inferences.github.io/short-course/)

## Nodal types

::: panel-tabset

```{r, include= FALSE}
library(CausalQueries)
```

# No parents

The type simply indicates the value of the node

```{r}
make_model("X -> Y") |>
  interpret_type(nodes = "X")
```

e.g. type X is either 0 or it is 1!


# One parent

The type indicates the value of the node give the two different possible values of the parents. There are four ($2^2$) possible types!

```{r}
make_model("X -> Y") |>
  interpret_type(nodes = "Y")
```

e.g. type "01": Y = 0 if X = 0, Y = 1 if X = 1

# Two parents

The type indicates the value of the node give the two different possible values for each of  the two parents. Since there are 4 possible combinations of parents there are sixteen ($2^4$) possible types!

```{r}
make_model("X -> Y <- Z") |>
  interpret_type(nodes = "Y")
```

e.g. type "0001": Y = 1 only if X = 1 and Z = 1


# Three parents

The type indicates the value of the node give the two different possible values for each of  the three parents. Since there are 8 possible combinations of parents there are 256 ($2^8$) possible types!

```{r}
make_model("A -> Y <- B; C->Y") |>
  interpret_type(nodes = "Y")
```

e.g. type "10000001": Y = 1 only if A = 0, B= 0 , C = 0 or A = 1, B = 1 , C = 1

Beyond 3 parents things really explode since there are > 60000 types for a single node!

# Note on ordering

The order in which nodes appear in the indexing depends on the their "generation" in the causal model. In this example C actually precedes B and so precedes it in the indexing:

```{r}
make_model("A -> Y <- B <- C; C->Y") |>
  interpret_type(nodes = "Y")
```

:::
