---
title: "solution"
format: 
  html:
    embed-resources: true
    code-fold: true
    number-sections: true
---

We print out the inference table and then provide a helper function to let you quickly see the conditional inferences. 

You can also see the solutions in [Chapter 12](https://integrated-inferences.github.io/book/12-clue-selection.html) or by using the [shiny](https://shiny2.wzb.eu/ipi/process_tracing/).

# Set up and helper function

```{r, include = TRUE, message = FALSE, warning = FALSE}
library(CausalQueries)
library(knitr)
library(tidyverse)

data_consistent <- function(my_table, seen = c("S1", "Y0")) {
  for(j in seen){
    my_table <- 
    my_table[c(names(my_table)[grepl(j,  names(my_table))], "in_query",   "priors")]
  }

  my_table <- my_table |> filter(
    my_table |> select(-in_query, -priors) |> apply(1, sum) > 0)

  my_table |> select(-in_query, -priors)
  
  pDQ <- my_table |> filter(in_query) |> pull(priors) |> sum()
  pD  <- my_table |> pull(priors) |> sum()
  print(c("prob of Data and Query:", pDQ))
  print(c("prob of Data:", pD))
  print(c("prob Query | Data:", pDQ/pD))
  
  print(my_table)

}  

```

# Model parameters

Check that the model has the parameters we want!

```{r}

model <- 
  make_model("S -> C -> Y <- R <- X; X -> C -> R") |>
  set_restrictions(labels = list(C = c('1110', '1111'), 
                                 R = c('0000', '0001'), 
                                 Y = '0001'),
                 keep = TRUE)

query <- "Y[X=1] > Y[X=0]"

model$parameters_df |> kable()

my_table <-
  model  |>
  grab(what = "ambiguities_matrix") |>
  data.frame() |>
  mutate(
    in_query = get_query_types(model, query)$types,
    priors   = CausalQueries:::get_type_prob(model)) 



```


# Look up table for conditional inferences

```{r}
my_table |> kable()
```

# Examples of conditional inferences 

(using a helper funciton to display the subset of the data we care about)

An inference:


## I observe $S=0, Y=0$ only:

```{r}
  data_consistent(my_table, seen = c("S0", "Y0"))
```



## I observe $S = 1, X = 1, C = 0$

```{r}

data_consistent(my_table, seen = c("S1", "X1", "C0"))
```
